---
title: "Lab 3A Starting with the Data"
author: "Wasique Hasan"
date: "09/20/2020"
output: html_document
chunk_output_type: console
---

library(knitr)
library(markdown)
library("tidyverse")

``` {r, eval = FALSE}
download.file(url="https://ndownloader.figshare.com/files/2292169",
destfile = "data/portal_data_joined.csv")
```

                
``` {r} 
surveys <- read.csv("data/portal_data_joined.csv")
## Start and end
head(surveys)
tail(surveys)
View(surveys)

## Additional information
dim(surveys)
nrow(surveys)
ncol(surveys)
names(surveys)
rownames(surveys)

## Summary
str(surveys)
summary(surveys)
```

## Challenge question 1\
Based on the output of str(surveys), can you answer the following questions?\

What is the class of the object surveys?\
Answer: The class of surveys is data frame.\
How many rows and how many columns are in this object?\
Answer: There are 34786 rows and 13 columns.\
How many species have been recorded during these surveys?\
Answer: 48 species.

## Challenge question 2
Create a data.frame (surveys_200) containing only the data in row 200 of the surveys dataset.\

Notice how nrow() gave you the number of rows in a data.frame?\

Use that number to pull out just that last row in the data frame.\
Compare that with what you see as the last row using tail() to make sure it’s meeting expectations.\
Pull out that last row using nrow() instead of the row number.\
Create a new data frame (surveys_last) from that last row.\
Use nrow() to extract the row that is in the middle of the data frame.\
Store the content of this row in an object named surveys_middle.\
Combine nrow() with the - notation above to reproduce the behavior of head(surveys), keeping just the first through 6th rows of the surveys dataset.\
```{r}
surveys_200 <- surveys[1:200, ]
nrow(surveys_200)
tail(surveys_200)
surveys_200[nrow(surveys_200), ]
surveys_last <- surveys_200[nrow(surveys_200), ]
surveys_200[nrow(surveys_200)/2, ]
surveys_middle <-surveys_200[nrow(surveys_200)/2, ]
surveys_200[-(7:nrow(surveys_200)), ]
```

```{r}
sex <- factor(c("male", "female", "female", "male"))
levels(sex)
nlevels(sex)
sex # current order
sex <- factor(sex, levels = c("male", "female"))
sex # after re-ordering
as.character(sex)
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990))
as.numeric(year_fct)               # Wrong! And there is no warning...
as.numeric(as.character(year_fct)) # Works...
as.numeric(levels(year_fct))[year_fct]    # The recommended way.
## bar plot of the number of females and males captured during the experiment:
plot(as.factor(surveys$sex))
```

```{r}
sex <- factor(surveys$sex)
head(sex)
levels(sex)
levels(sex)[1] <- "undetermined"
levels(sex)
head(sex)
```
## Challenge question 3
Rename “F” and “M” to “female” and “male” respectively.
Now that we have renamed the factor level to “undetermined”, can you recreate the barplot such that “undetermined” is last (after “male”)?
```{r}
levels(sex)[2] <- "Female"
levels(sex)[3] <- "Male"
sex<- factor(sex, levels=c("Female", "Male", "undetermined"))
plot(sex)
```

```{r}
## Compare the difference between our data read as `factor` vs `character`.
surveys <- read.csv("data/portal_data_joined.csv", stringsAsFactors = TRUE)
str(surveys)
surveys <- read.csv("data/portal_data_joined.csv", stringsAsFactors = FALSE)
str(surveys)
## Convert the column "plot_type" into a factor
surveys$plot_type <- factor(surveys$plot_type)
```
## Challenge question 4
We have seen how data frames are created when using read.csv(), but they can also be created by hand with the data.frame() function. There are a few mistakes in this hand-crafted data.frame. Can you spot and fix them? Don’t hesitate to experiment!

Corrected version:\
```{r}
animal_data <- data.frame(
          animal = c("dog", "cat", "sea cucumber", "sea urchin"),
          feel = c("furry", "furry", "squishy", "spiny"),
          weight = c(45, 8, 1.1, 0.8)
          )
```

Can you predict the class for each of the columns in the following example? Check your guesses using str(country_climate):\
Answer: has_kangaroo is numeric, all the others are factors.
Are they what you expected? Why? Why not?\
Answer: Yes, they are what I expected.
What would have been different if we had added stringsAsFactors = FALSE when creating the data frame?\
Answer: There would be character vectors instead of factors. 
What would you need to change to ensure that each column had the accurate data type?\
Answer: Removing quotes in temperature and northern_hemisphere and replacing the 1 with TRUE in has_kangaroo. 
```{r}
country_climate <- data.frame(
       country = c("Canada", "Panama", "South Africa", "Australia"),
       climate = c("cold", "hot", "temperate", "hot/temperate"),
       temperature = c(10, 30, 18, "15"),
       northern_hemisphere = c(TRUE, TRUE, FALSE, "FALSE"),
       has_kangaroo = c(FALSE, FALSE, FALSE, 1)
       )
str(country_climate)
```

```{r}
str(surveys)
library("lubridate")
my_date <- ymd("2015-01-01")
str(my_date)
# sep indicates the character to use to separate each component
my_date <- ymd(paste("2015", "1", "1", sep = "-")) 
str(my_date)
paste(surveys$year, surveys$month, surveys$day, sep = "-")
ymd(paste(surveys$year, surveys$month, surveys$day, sep = "-"))
surveys$date <- ymd(paste(surveys$year, surveys$month, surveys$day, sep = "-"))
str(surveys) 
summary(surveys$date)
missing_dates <- surveys[is.na(surveys$date), c("year", "month", "day")]
head(missing_dates)
```
